

Now that we know how to manipulate, combine and save plots, let's check some other graphs that can be made using `ggplot2`. In particular, we will present those that we have been using in our computer labs.

## Histogram and bar plot

### Histogram

Imagine that we want to see how the price of the diamonds is distributed, i.e.: what ranges of prices occur more frequently in our data sample. For that, we can make a histogram with the following code.

```{r out.width = '50%', fig.pos = 'h', fig.align = "center", fig.cap="\\label{fig:hist1}A histogram with ggplot2."}
plotHist1 <- ggplot(dsmall)
plotHist1 <- plotHist1 + geom_histogram(aes(price, 
                                            y = ..density.. # remove for absolute values
                                            ),
                                        color = "black",
                                        fill  = "lightblue",
                                        bins  = 10 # no need to establish this value
                                       )
plotHist1 <- plotHist1 + labs(title = "Histogram")
print(plotHist1)
```

Notice that the part related with `aes`, instead of being defined inside `geom_histogram`, it could have been defined inside the function `ggplot`, as we did in the previous examples. Depending on the situation, one option will be better than the other. Practice will tell you. 

### Stacked Histogram

Sometimes it might happen that, in addition to see how a variable is distributed, we also want to check, for each bar of the plot, which elements are contributing and how, i.e.: we want a stacked histogram. As an example, imagine that in our previous plot we wanted to see, for each bar in the histogram, what is the contribution per type of cut. The code for that is:  

```{r out.width = '50%', fig.pos = 'h', fig.align = "center", fig.cap="\\label{fig:hist2}A stacked histogram with ggplot2."}
plotHist2 <- ggplot(dsmall)
plotHist2 <- plotHist2 + geom_histogram(aes(price, fill = cut),
                                      color = "black",
                                      bins = 10 # no need to stablish this value
                                     )
plotHist2 <- plotHist2 + scale_fill_viridis_d(option = "magma", alpha = 0.8)
plotHist2 <- plotHist2 + labs(title = "Stacked histogram")
print(plotHist2)
```

### Bar plot

In case that we wanted, for example, to visualize the proportion of diamonds in the data set per type of cut, we would use a bar plot instead of a histogram because variable `cut`, opposite to variable `price`, is discrete.

```{r out.width = '50%', fig.pos = 'h', fig.align = "center", fig.cap="\\label{fig:bar1}A stacked bar plot with ggplot2."}
plotBar <- ggplot(dsmall)
plotBar <- plotBar + geom_bar(aes(cut, fill = color), 
                              width = 0.5) # width of the bars
plotBar <- plotBar + scale_fill_viridis_d()
plotBar <- plotBar + labs(title = "Stacked bar plot")

# Let's also rotate the x-labels
plotBar <- plotBar + theme(axis.text.x = element_text(angle = 65, vjust = 0.6))
print(plotBar)
```

Notice that we have also split the composition of each bar according to the variable `color` in the data set.


## Density plot

In the histogram of Figure \ref{fig:hist1}, the distribution of the variable `price` seemed to be exponential, but can we assume that? During the course, we have studied different non-parametrical tests to answer that kind of questions. However, to have a graphical intuition about what answer we can expect, we could just add the density line of the corresponding distribution to the histogram and see how well it fits the data.

So, as an example, let's add to the first histogram an exponential and a normal density line. The required parameters $\lambda$, $\mu$ and $\sigma$ will be estimated from the data. Also, we will add the corresponding equations of those distributions using `annotate`, a function which allows us to add any kind of text to a plot.

```{r out.width = '50%', fig.pos = 'h', fig.align = "center", fig.cap="\\label{fig:density}Adding density lines to a plot."}
plotDensity <- plotHist1 + labs(title = "Histogram and density line")

# Exponential distribution
plotDensity <- plotDensity + stat_function(fun   = dexp, 
                                           args  = list(rate = 1/mean(dsmall$price)), 
                                           color = "darkred", size = 0.7)

plotDensity <- plotDensity + annotate("text", x = 5000, y = 0.00020, 
                                      parse = TRUE, size  = 4, 
                                      label = "y == lambda*e^{-lambda*x}", 
                                      color = "darkred")

# Normal distribution
plotDensity <- plotDensity + stat_function(fun  = dnorm, 
                                           args = list(mean = mean(dsmall$price), 
                                                       sd   = sd(dsmall$price)), 
                                           color = "darkgreen", size = 0.7)

plotDensity <- plotDensity + annotate("text", x = 10000, y = 0.00010, 
                                      parse = TRUE, size = 4, 
                                      label = "y==frac(1, sqrt(2*pi))*e^{-x^2/2}", 
                                      color = "darkgreen")

print(plotDensity)
```


As you can see, none of the distributions really fits the data. For the normal distribution that was expected, but, what about the exponential one? Although the shape of the data fits the density line, the empirical frequencies do not match with the theoretical ones (at least in the first quantiles). However, as mentioned in other parts of this document, our conclusions should be based on maths, not in plots.

\begin{shaded}
{\color{red}{\textbf{Note:}}}
Check on the Internet what other things, besides ``text'', can be added using \texttt{annotate}.
\end{shaded}



## Regression line

Now let's learn how to make a scatter plot containing the corresponding regression line using `ggplot2`. For that, consider the example that we used in the first plot that we made in this document (Figure \ref{fig:ExamplePlot}). In the mentioned example we made: "for each type of cut, a scatter plot to visualize the price of the diamonds according to their weight." So, let's make that plot, but this time, adding the corresponding regression line and changing the shape and color of the points whose price is above 15K.

```{r out.width = '80%', fig.pos = 'h', fig.align = "center", fig.cap="\\label{fig:OLS}Plotting OLS with ggplot2."}
plotReg1 <- ggplot(dsmall)
plotReg1 <- plotReg1 + facet_wrap(~ cut, ncol = 3)
plotReg1 <- plotReg1 + geom_point(aes(x = carat, y = price, 
                                      shape = price > 15000, # condition for the shape
                                      color = price > 15000), 
                                  alpha = 0.5)
plotReg1 <- plotReg1 + scale_shape_manual(values = c(16,8)) # choose the shape
plotReg1 <- plotReg1 + scale_color_manual(values = c("black", "red")) # choose the color
plotReg1 <- plotReg1 + geom_smooth(aes(x = carat, y = price), method = lm, se = FALSE)
plotReg1 <- plotReg1 + theme(legend.position = "none")
print(plotReg1)
```

\begin{shaded}
{\color{red}{\textbf{Note:}}}
If `se = FALSE` is removed, the plot will show the confidence intervals for $\hat{y}$. Additionally, it is possible to change the color of the line adding \texttt{col = ``nameOfTheColor''} inside \texttt{geom\_smooth}.
\end{shaded}


### Multiple regression lines

Sometimes, we might want to add, in the same plot, a different regression line according to a categorical variable. As an example of that, check the following code: 

```{r out.width = '60%', fig.pos = 'h', fig.align = "center", fig.cap="\\label{fig:OLS2}Multiple regression lines."}
# We use a smaller data set (for clarity) considering 3 types of cut
dsmall2  <- dsmall[dsmall$cut %in% c("Fair", "Good", "Very Good"), ]

# The plot
plotReg2 <- ggplot(dsmall2) + labs(title = "Multiple lines")
plotReg2 <- plotReg2 + geom_point(aes(x = carat, y = price, color = cut, shape = cut), 
                                  alpha = 0.5)
plotReg2 <- plotReg2 + geom_smooth(aes(x = carat, y = price, color = cut), 
                                   method = lm, se = FALSE)
plotReg2 <- plotReg2 + scale_color_viridis_d(option = "inferno", 
                                             end = 0.75) # to not use the yellow color
print(plotReg2)
```




## Laying out multiple plots on the same frame

There are occasions in which we might want to make several plots on a unique frame to simultaneously expose different ideas about the same data set. In this section we will see a brief introduction to that. However, before doing it, let's change the text size appearing in the three plots (Figures \ref{fig:hist2}, \ref{fig:density}, \ref{fig:OLS2}) that we will be using in our example. This is to avoid the final plot not looking good (e.g.: having an overlap of the values in the axes).


```{r}
plotHist2 <- plotHist2 + theme(plot.title  = element_text(size = 10),
                               axis.text.x = element_text(size = 6),
                               axis.text.y = element_text(size = 6),
                               text = element_text(size = 7)
                              )

plotDensity <- plotDensity + theme(plot.title   = element_text(size = 10),
                                   legend.title = element_text(size = 7),
                                   axis.text.x  = element_text(size = 6),
                                   axis.text.y  = element_text(size = 6),
                                   text = element_text(size = 7)
                                  )

plotReg2 <- plotReg2 + theme(plot.title   = element_text(size = 10),
                             legend.title = element_text(size = 7),
                             axis.text.x  = element_text(size = 6),
                             axis.text.y  = element_text(size = 6),
                             text = element_text(size = 7)
                            )
```


Now, imagine that we want to represent the three previous plots on the same frame. For that, we need to provide how we want to divide the frame (e.g.: 2 columns and 2 rows) and put the plots on it. **Additionally**, it is also important to label each of the plots so we can properly reference them in the text afterwards. Check the following code to see how to achieve this.

```{r message = FALSE, warning = FALSE, out.width = '80%', fig.align = "center", fig.pos = 'h', fig.cap="\\label{fig:multiple1}Multiple plots in the same frame."}
library(cowplot)
multiplePlots <- plot_grid(plotHist2, plotDensity, plotReg2, 
                           labels = "AUTO",  # use "auto" for lowercase letters
                           label_size = 8,
                           ncol = 2, nrow = 2)

title <- ggdraw() + draw_label("General Title", fontface = 'bold')

plot_grid(title, multiplePlots, ncol = 1, 
          rel_heights = c(0.1, 1)) # rel_heights values control title margins
```

\begin{shaded}
{\color{red}{\textbf{Note:}}}
Try to remove \texttt{rel\_heights} to see what happens.
\end{shaded}


Now check the same example, but when making one of the plots bigger than the other two.

```{r warning = FALSE, out.width = '80%', fig.pos = 'h', fig.align = "center", fig.cap="\\label{fig:multiple2}Multiple plots in the same frame II."}
multiplePlots2 <- plot_grid(plotHist2, 
                            plot_grid(plotDensity, plotReg2, 
                                      ncol = 2, nrow = 1, 
                                      labels = c("(B)", "(C)"), label_size = 8), 
                            labels = c("(A)"), label_size = 8,
                            ncol = 1, nrow = 2)

plot_grid(title, multiplePlots2, ncol = 1, rel_heights = c(0.1, 1)) 
```

